---
title: "Assignment 2: USGS Stream"
author: "Natasha Atkins"
format: pdf
editor: visual
---

## Week 3: Git and Stream Temperature

## Downloading the USGS Data

```{r}
#install.packages("dataRetrieval") 
library(dataRetrieval) 
library(tidyverse) 
library(here)  

#get usgs data 
siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

#save data as a csv
write.csv(temp_data, here("Data", "usgs_stream_data_assn2.csv"), row.names = FALSE)
```

### Load data in instead – can use after initial download from server

```{r}
temp_data_dwnld <- read.csv(here("Data","usgs_stream_data_assn2.csv"))
```

## Data processing

```{r}
# rename columns for clarity
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```

### Calculate summary stats

```{r}
#fixing column names error
ncol(temp_data)
colnames(temp_data)

colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C", "quality_cd", "col6", "col7")

# calculate mean, min, max temperature per site
site_means <- temp_data |>
  group_by(site_no) |>
  summarise(
    mean_temp_C = mean(temp_C, na.rm = TRUE),
    n_obs       = n(),
    min = min(temp_C, na.rm = TRUE),
    max = max(temp_C, na.rm = TRUE)
  )

print(site_means)
```

## Natasha's Analysis

### First, branch my individual analysis

In the terminal type: Git branch Atkins – to establish the new branch. Then type: Git checkout Atkins to go to that branch.

After branching, begin conducting analysis

```{r}
#how is month stored
str(temp_data$Date)
# filter for June temperatures -- addressing the Date format
library(lubridate)
library(dplyr)

temp_data <- temp_data |>
  mutate(month = month(Date),
         year = year(Date))  # extract numeric month and year

june_data <- temp_data |>
  filter(month == 6)

#filter for june temperatures (based on provided code)
#june_data <- temp_data |>
  #filter(month == "6")

# fit a GLM model to see if there is a trend in June temperatures over time, separated by site
june_model <- glm(temp_C ~ year*site_no, data = june_data)
summary(june_model)
```

Visualizing the glm

```{r}
ggplot(june_data, aes(x = year, y = temp_C, color = as.factor(site_no))) +
  #geom_point(alpha = 0.3) + # Raw data points
  geom_smooth(method = "glm", formula = y ~ x, size = 1) + # Model fits
  labs(x = "Year", y = "Temperature (C)", color = "Site") +
  theme_minimal()

```
