---
title: "Homework 3"
author: "Natasha Atkins"
format: pdf
editor: visual
---

# Exploring Looping with Air Quality Data

## Part 1: Build the dataset

```{r}
#load libraries
library(tidyverse)
library(janitor)
library(ggplot2)

#create pm2.5 data
pm25 = runif(min=0, max=50, n=365)
total = 0
nvalues = 365
pm25

#days
day = (1:365)

#seasons
season = ifelse(day <= 90, "winter",
         ifelse(day <= 180, "spring",
         ifelse(day <= 270, "summer", "fall")))
season = as.factor(season)

#risk
risk = rep(NA, 365)

#build data frame
aq_dataframe = data.frame(day=day, pm25=pm25, season=season, risk=risk )
head(aq_dataframe)
summary(aq_dataframe)
```

## Part 2: Classify risk with a for loop

```{r}
#classify the thresholds

#good_aq <= 12
#moderate_aq > 12 & <= 35.4
#unhealthy_aq > 35.4

ndays = 365
#initializing the variables
max_unhealthy_streak = 0
current_streak = 0

#the beginning of the loop says to look through the rows (days) and for every pm value in that row determine if it meets the if condition
#if the values are classified as good or moderate then they don't contribute to theunhealhty streak, so current_streak = 0

for (i in 1:ndays){
    if(aq_dataframe[i,"pm25"] <= 12) {
      aq_dataframe[i, "risk"] = "good"
      current_streak = 0
    } else {
        if (aq_dataframe[i,"pm25"] <= 35.4){
        aq_dataframe[i, "risk"] = "moderate"
        current_streak = 0
        } else {
          aq_dataframe[i, "risk"] = "unhealthy"
          current_streak = current_streak +1
          }
    } 
      if (current_streak > max_unhealthy_streak){
      max_unhealthy_streak = current_streak
      }
} 
print(max_unhealthy_streak)
```

## Part 3: Seasonal summary with a for loop

```{r}
# initialize vector
seasonal_unhealthy = c(winter = 0, spring = 0, summer = 0, fall = 0)

# iterate by looking at each row for each day
for (i in 1:ndays){
  # check for unhealthy day --> and if the day is unhealhty check the season
  if (aq_dataframe[i, "risk"] == "unhealthy"){
    if (aq_dataframe[i, "season"] == "winter"){
      # count the day toward the correct season
      seasonal_unhealthy[1] = seasonal_unhealthy[1]+1 
    } else if (aq_dataframe[i, "season"] == "spring"){
      seasonal_unhealthy[2] = seasonal_unhealthy[2]+1
    } else if (aq_dataframe[i, "season"] == "summer"){
      seasonal_unhealthy[3] = seasonal_unhealthy[3]+1
    } else {
      seasonal_unhealthy[4] = seasonal_unhealthy[4]+1
    }
  }
}

print(seasonal_unhealthy)
```

## Part 4: Visualize the data with plots

```{r}
#time series of pm2.5 values
aq_plot <- ggplot(aq_dataframe, aes(x = day, y = pm25, color = season)) +
  geom_line() +
  theme_bw() +
  labs(
    title = sprintf("PM2.5 Levels Across the Year with the Maximum Number of Unhealthy Days Being %d days", max_unhealthy_streak),
    x = "Day",
    y = "PM2.5 Level"
  )
aq_plot

```

```{r}
# bar chart of unhealthy days per season
# build a data frame to access the number of unhealthy days for the plot
seasonal_unhealthy_df <- data.frame(
  season = names(seasonal_unhealthy),
  count = as.numeric(seasonal_unhealthy)
)

ggplot(seasonal_unhealthy_df, aes(x = season, y = count, fill = season)) +
  geom_col() +
  theme_bw() +
  labs(
    title = sprintf("Unhealthy Days by Season with there being %d Days in Winter, %d Days in Spring, %d Days in Summer, and %d Days in Fall",
                    seasonal_unhealthy["winter"], seasonal_unhealthy["spring"],
                    seasonal_unhealthy["summer"], seasonal_unhealthy["fall"]),
    x = "Season",
    y = "Number of Unhealthy Days"
  )
```

## Summary – Interpretation for Managers

The major patterns revealed from this analysis are that the longest streak of unhealthy days was 4 days in a row with unhealthy air quality (PM2.5 levels above 35.4 ug/m\^3). This unhealthy streak occurred during the winter on days 54-57 of the year. The season with the highest number of unhealthy days total was the summer with 33 unhealthy days. The season with the lowest was spring with 23. While all seasons had relatively similar amount of unhealthy days (as is expected with the uniform distribution we set to generate the data), spring having 10 less days than summer may indicate that scheduling outdoor activities, such as construction, land management, recreational events, might be safer to do in the spring than in the summer.
