---
title: "Assignment 2"
author: "Sydney Royal"
format: html
editor: visual
---

## Assignment Description

Working together create a new repository that you will share

- add a quarto file, with nice documentation, that downloads the data (call it something informative)

- also store the data in the repository (e.g. as a csv or RData file) so that you can work with it without having to download it again

- in your quarto file, add the option to load the data instead of accessing from NWIS server

- add a code chunk that plots the data (e.g. using ggplot)

- add a code chunk that calculates some summary statistics (e.g. mean, max, min temperature for each site)

Make sure you push to *GitHub*.

Now imagine you are working on a project, where you are looking trends in water temperature you want to figure out if a) there is a trend and b) if the slope (trend) is positive (increasing) or negative (decreasing) and c) if there are differences between sites

Have each partner focus on trends for a different month

Here’s some example code for how you might compute the trend of temperatures in a given month, the slope of the GLM model will tell you what the trend is (how much temperature increases/decreases per year)

- Estimate for *year* tells you slope for first site

- Estimate for *year:site_no14211010*, *year:site_no14211720* tells you how much greater (if positive) slope is for this site relative to first site

- Estimates for *Intercept* and *site_no14211010*, *site_no14211010* set the baseline temperature to make regression work. You won’t use these. (An easier way to look at differences in mean temperature across sites is to calculate means for each)

## Install Needed Packages

```{r}

#install.packages("dataRetrieval")
library(dataRetrieval)
library(tidyverse)
library(here)


```

## Download the Data from the server

```{r}

siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

#temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)

# NAT ALREADY SAVED IT IN THE REPO
```

## Download the Data from the repo instead

```{r}

temp_data_dwnld <- read.csv(here("Data","usgs_stream_data_assn2.csv"))

```

## Clean it up and Plot it

```{r}
# rename columns for clarity
colnames(temp_data_dwnld) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data_dwnld$month = month(temp_data_dwnld$Date, label=TRUE)
temp_data_dwnld$year = year(temp_data_dwnld$Date)

# check by plotting
ggplot(temp_data_dwnld, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```

## Computing Trends (May)

```{r}

# filter for July temperatures
may_data <- subset(temp_data_dwnld, month == "May")
# fit a GLM model to see if there is a trend in July temperatures over time, separated by site
may_model <- glm(temp_C ~ year*site_no, data = may_data)
summary(may_model)


```

## Calculate Averages Across Sites

```{r}

# AVERAGE TEMPERATURE ACROSS ALL SITES
temp_mean<- mean(temp_data_dwnld$temp_C)


# FIX COLUMN NAMES
colnames(temp_data_dwnld)<- c("agency_cd","site_no","Date","temp_C","quality_cd","month","year")

# AVERAGE GROUPED BY SITES (TABLE)
sitemeans<- temp_data_dwnld|>
  group_by(site_no)|>
  summarize(mean_temp = mean(temp_C, na.rm= TRUE))

sitemeans


```
